---
title: "Gillnet DFA exploration"
output:
  pdf_document: default
  html_notebook: default
date: "June 01, 2017"
---

This is for the top sharks, turtles, mammals, and birds caught in gillnets. 23 species: 10 sharks, 5 mammals, 4 turtles, 4 birds. 

Plot z-scored trends for all gillnets
```{r, z-scored top 10 trends, echo = FALSE, fig.align='center'}

# load ("Z:/Julia/Peru/Peru_data_exploration/Output/DFA_top_gn_best_fit.RData")
# load ("Z:/Julia/Peru/Peru_data_exploration/Output/DFA_top_gn_best_model.RData")


gn.top.mn.ts <- read.csv("../Data_tables/Top_gn_DFA_ts.csv")
dat.z.gn <- as.matrix (gn.top.mn.ts[,-1]) # remove spp column

Sigma.gn = sqrt(apply(dat.z.gn, 1, var, na.rm=TRUE)) 
y.bar.gn = apply(dat.z.gn, 1, mean, na.rm=TRUE) 
z.sco.gn = (dat.z.gn - y.bar.gn) * (1/Sigma.gn)

years = colnames (z.sco.gn)

spp.gn <- gn.top.mn.ts$Species

#pdf (file = "../Output/GN_CPUE_Zscored_trends.pdf")
par(mfcol=c(5,2), mar=c(3,4,1.5,0.5), oma=c(0.4,1,1,1))
for(i in 1:length(spp.gn)){
  #plot(years, z.sco[i,], xlab="",ylab="CPUE index", bty="L", pch=16, col="blue", type="b")
  plot(z.sco.gn[i,], xlab="",ylab="CPUE index", bty="L", pch=16, col="blue", xaxt = "n", type="b")
  axis(1, at = seq (0,11, by = 5), labels = seq(2005, 2016, by = 5) )# can't get this to work
  title(spp.gn[i])
}
#dev.off()
```

Plot factor loadings
```{r, plot factor loadings, echo = FALSE, fig.align= 'center'}
load ("../Output/DFA_top_gn_best_fit.RData")
load ("../Output/DFA_top_gn_best_model.RData")

# determine loadings and trends

H.inv.gn = varimax (coef (best.fit.top.gn, type = "matrix") $ Z) $ rotmat

# rotate factor loadings
Z.rot.gn = coef (best.fit.top.gn, type = "matrix") $ Z %*% H.inv.gn

# rotate trends
trends.rot.gn = solve (H.inv.gn) %*% best.fit.top.gn$states

# Plot
N.ts.gn = 23
minZ = 0.05
ylims = c(-1.1*max(abs(Z.rot.gn)), 1.1*max(abs(Z.rot.gn)))

pdf (file = "../Output/GN_CPUE_factor_loadings.pdf")
for (i in 1:best.model.top.gn$m) {
plot(c (1:N.ts.gn)[abs (Z.rot.gn[,i]) > minZ], 
     as.vector(Z.rot.gn[abs(Z.rot.gn[,i]) > minZ, i]),
       type = "h", lwd = 2, xlab = "", ylab = "", xaxt = "n", ylim = ylims, xlim = c(0, N.ts.gn + 1))
  for(j in 1:N.ts.gn) {
    if(Z.rot.gn[j,i] > minZ) {text(j, -0.05, spp.gn[j], srt = 90, adj = 1, cex = 0.9)}
    if(Z.rot.gn[j,i] < -minZ) {text(j, 0.05, spp.gn[j], srt = 90, adj = 0, cex = 0.9)}
    abline(h = 0, lwd = 1, col = "gray")
  } # end j loop
  mtext (paste ("Factor loadings on trend", i, sep = " "), side = 3, line = 0.5)
} # end i loop
dev.off()

```

Plot trends
```{r, plot trends, echo = FALSE, fig.align = "center"}

ts.trends.gn <- t(trends.rot.gn)

pdf (file = "../Output/GN_CPUE_DFA_trends.pdf")
for (i in 1:dim (ts.trends.gn)[2]) {
  # set up plot area
  plot ( ts.trends.gn[,i],
        ylim = c(-1.1, 1.1) * max (abs (ts.trends.gn)),
        type = "n", lwd = 2, bty = "L",
        xlab = "", ylab = "", yaxt="n", xaxt = "n")
  # Draw zero line
  abline (h = 0, col = "gray")
  # Plot trend line
  par (new = TRUE)
  plot ( ts.trends.gn[,i],
        ylim = c(-1.1, 1.1) * max (abs (ts.trends.gn)),
        type = "l", lwd = 2, bty = "L",
        xlab = "", ylab = "", xaxt = "n")
  # Add panel labels
  mtext (paste ("Trend", i, sep = " "), side = 3, line = -2, cex = 1.5, font = 2)
  axis(1, at = seq (0,11, by = 2), labels = seq(2005, 2016, by = 2))
} # end i loop
dev.off()

```

Plot fits
```{r, plot model fits gn}
# TT = length of time series
TT <- 12
par.mat.gn <- coef (best.fit.top.gn, type = "matrix")
fit.b.gn <- par.mat.gn$Z %*% best.fit.top.gn$states + matrix (par.mat.gn$A, nrow = N.ts.gn, ncol = TT)

pdf (file = "../Output/GN_CPUE_DFA_fits.pdf")
par (mfcol = c (3, 2), mar = c(3, 4, 1.5, 0.5), oma = c(0.4, 1, 1, 1))
for (i in 1:length (spp.gn)) {
plot(dat.z.gn[i,], xlab = "", ylab = "abundance index", bty = "L", xaxt = "n", ylim = c (min( fit.b.gn[i,]), 2* max (dat.z.gn[i,])), 
     pch = 16, col = "blue")
 axis(1, at = seq (0,11, by = 5), labels = seq(2005, 2016, by = 5))
 lines (fit.b.gn[i,], lwd = 2)
 title (spp.gn[i])
}
dev.off()

```

Add covariates
```{r, covariates, echo = FALSE}
# ONI: http://ggweather.com/enso/oni.htm
# has to be numeric. codes for neutral is zero, weak is +-1, medium is +-2, VS is +-4
oni.num <- c(0,1,-2,0,2, -2, -1, 0,0,0,4, -1)

pdo <- c(0.29, 0.06, -1.16, -0.97, -0.47, -0.62, -1.44, -1.23, -0.60, 1.13, 1.57, 1.31)

model.list.gn <- list(m = 2, R = "equalvarcov") 
cntl.list <-  list(minit = 200, maxit = 5000, allow.degen = FALSE) 

kemz.gn.oni <-  MARSS(dat.z.gn, model = model.list.gn, z.score = TRUE, form = "dfa", control = cntl.list, covariates = oni.num)

kemz.gn.pdo  <- MARSS(dat.z.gn, model = model.list.gn, z.score = TRUE, form = "dfa", control = cntl.list, covariates = pdo)

kemz.gn.both  <- MARSS(dat.z.gn, model = model.list.gn, z.score = TRUE, form = "dfa", control = cntl.list, covariates = rbind(oni.num, pdo))

```

Is it better?

```{r, AIC, echo = FALSE}
print(cbind(model=c("no covars", "ONI", "PDO", "ONI & PDO"), AICc=round(c(best.fit.top.gn$AICc, kemz.gn.oni$AICc, kemz.gn.pdo$AICc, kemz.gn.both$AICc))), quote=FALSE)
```

Covariates do not improve the model. 


### Now with Salaverry


This is for the top sharks, turtles, mammals, and birds caught in gillnets JUST in Salaverry.16 species: 6 sharks, 4 mammals, 4 turtles, 2 birds. 

Plot z-scored trends for all gillnets
```{r, svry z-scored top 10 trends, echo = FALSE, fig.align='center'}

svry.top.mn.ts <- read.csv("../Data_tables/Top_svry_DFA_ts.csv")
dat.z.svry <- as.matrix (svry.top.mn.ts[,-1]) # remove spp column

Sigma.svry = sqrt(apply(dat.z.svry, 1, var, na.rm=TRUE)) 
y.bar.svry = apply(dat.z.svry, 1, mean, na.rm=TRUE) 
z.sco.svry = (dat.z.svry - y.bar.svry) * (1/Sigma.svry)

years = colnames (z.sco.svry)

spp.svry <- svry.top.mn.ts$Species

pdf (file = "../Output/svry_CPUE_Zscored_trends.pdf")
par(mfcol=c(5,2), mar=c(3,4,1.5,0.5), oma=c(0.4,1,1,1))
for(i in 1:length(spp.svry)){
  #plot(years, z.sco[i,], xlab="",ylab="CPUE index", bty="L", pch=16, col="blue", type="b")
  plot(z.sco.svry[i,], xlab="",ylab="CPUE index", bty="L", pch=16, col="blue", xaxt = "n", type="b")
  axis(1, at = seq (0,11, by = 5), labels = seq(2005, 2016, by = 5) )# can't get this to work
  title(spp.svry[i])
}
dev.off()
```

Plot factor loadings
```{r, plot svry factor loadings, echo = FALSE, fig.align= 'center'}
load ("../Output/DFA_top_svry_best_fit.RData")
load ("../Output/DFA_top_svry_best_model.RData")

H.inv.svry = varimax (coef (best.fit.top.svry, type = "matrix") $ Z) $ rotmat

# rotate factor loadings
Z.rot.svry = coef (best.fit.top.svry, type = "matrix") $ Z %*% H.inv.svry

# rotate trends
trends.rot.svry = solve (H.inv.svry) %*% best.fit.top.svry$states


# Plot
N.ts.svry = 16
minZ = 0.05
ylims = c(-1.1*max(abs(Z.rot.svry)), 1.1*max(abs(Z.rot.svry)))

pdf (file = "../Output/svry_CPUE_factor_loadings.pdf")
for (i in 1:best.model.top.svry$m) {
plot(c (1:N.ts.svry)[abs (Z.rot.svry[,i]) > minZ], 
     as.vector(Z.rot.svry[abs(Z.rot.svry[,i]) > minZ, i]),
       type = "h", lwd = 2, xlab = "", ylab = "", xaxt = "n", ylim = ylims, xlim = c(0, N.ts.svry + 1))
  for(j in 1:N.ts.svry) {
    if(Z.rot.svry[j,i] > minZ) {text(j, -0.05, spp.svry[j], srt = 90, adj = 1, cex = 0.9)}
    if(Z.rot.svry[j,i] < -minZ) {text(j, 0.05, spp.svry[j], srt = 90, adj = 0, cex = 0.9)}
    abline(h = 0, lwd = 1, col = "gray")
  } # end j loop
  mtext (paste ("Factor loadings on trend", i, sep = " "), side = 3, line = 0.5)
}
dev.off()

```

Plot trends
```{r, plot wvry trends, echo = FALSE, fig.align = "center"}

ts.trends.svry <- t(trends.rot.svry)

pdf (file = "../Output/svry_CPUE_DFA_trends.pdf")
for (i in 1:dim (ts.trends.svry)[2]) {
  # set up plot area
  plot ( ts.trends.svry[,i],
        ylim = c(-1.1, 1.1) * max (abs (ts.trends.svry)),
        type = "n", lwd = 2, bty = "L",
        xlab = "", ylab = "", yaxt="n", xaxt = "n")
  # Draw zero line
  abline (h = 0, col = "gray")
  # Plot trend line
  par (new = TRUE)
  plot ( ts.trends.svry[,i],
        ylim = c(-1.1, 1.1) * max (abs (ts.trends.svry)),
        type = "l", lwd = 2, bty = "L",
        xlab = "", ylab = "", xaxt = "n")
  # Add panel labels
   mtext (paste ("Trend", i, sep = " "), side = 3, line = -2, cex = 1.5, font = 2)
  axis(1, at = seq (0,11, by = 2), labels = seq(2005, 2016, by = 2))
}
dev.off()

```

Plot fits
```{r, plot model fits svry}
# TT = length of time series
TT <- 12
par.mat.svry <- coef (best.fit.top.svry, type = "matrix")
fit.b.svry <- par.mat.svry$Z %*% best.fit.top.svry$states + matrix (par.mat.svry$A, nrow = N.ts.svry, ncol = TT)

pdf (file = "../Output/svry_CPUE_DFA_fits.pdf")
par (mfcol = c (3, 2), mar = c(3, 4, 1.5, 0.5), oma = c(0.4, 1, 1, 1))
for (i in 1:length (spp.svry)) {
plot(dat.z.svry[i,], xlab = "", ylab = "abundance index", bty = "L", xaxt = "n", ylim = c (min( fit.b.svry[i,]), 2* max (dat.z.svry[i,])), 
     pch = 16, col = "blue")
 axis(1, at = seq (0,11, by = 5), labels = seq(2005, 2016, by = 5))
 lines (fit.b.svry[i,], lwd = 2)
 title (spp.svry[i])
}
dev.off()

```

Add covariates
```{r, svry covariates, echo = FALSE}

model.list.svry <- list(m = 2, R = "diagonal and equal") 
cntl.list <-  list(minit = 200, maxit = 5000, allow.degen = FALSE) 

kemz.svry.oni <-  MARSS(dat.z.svry, model = model.list.svry, z.score = TRUE, form = "dfa", control = cntl.list, covariates = oni.num)

kemz.svry.pdo  <- MARSS(dat.z.svry, model = model.list.svry, z.score = TRUE, form = "dfa", control = cntl.list, covariates = pdo)

kemz.svry.both  <- MARSS(dat.z.svry, model = model.list.svry, z.score = TRUE, form = "dfa", control = cntl.list, covariates = rbind(oni.num, pdo))

```

Is it better?

```{r, AIC, echo = FALSE}
print(cbind(model=c("no covars", "ONI", "PDO", "ONI & PDO"), AICc=round(c(best.fit.top.svry$AICc, kemz.svry.oni$AICc, kemz.svry.pdo$AICc, kemz.svry.both$AICc))), quote=FALSE)
```
nope.